---
title: "NMFS Section 7 Storyboard"
author: "Michael Evans"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    ##vertical_layout: scroll
    #orientation: rows
    #css: custom.css
    social: menu
    df_print: paged
    navbar:
    - {title: "", icon: "fa-question-circle fa-lg", align: right, href: "mailto:esa@defenders.org?subject=NMFS section 7"}
    - {title: "", icon: "fa-github fa-lg", align: right, href: "https://github.com/mjevans26/NMFSs7"}
    - {title: "CCI", align: right, href: "https://cci-dev.org"}
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cooccur)
library(dplyr)
library(flexdashboard)
library(plotly)
library(reshape2)
library(shiny)
library(vegan)
library(viridis)

good_data <- readRDS(file = 'data/gooddata.rds')
species_x_category <- readRDS(file = "data/speciesxcat.rds")
df <- reactiveVal(good_data)

select_taxa <- selectInput('taxa', 'Select Taxon',
                           sort(c("", unique(good_data$Common.Name))), selected = "")
select_category <- selectInput('category', 'Select Work Type',
                               sort(c("", unique(good_data$Category))), selected = "")
select_agency <- selectInput('agency', 'Select Federal Agency',
                             sort(c("", unique(good_data$Agency))), selected = "")
run_update <- actionButton('update', 'Update')

observeEvent(input$update,
             {d <- filter(good_data,
                          grepl(input$category, Category),
                          grepl(input$taxa, Common.Name),
                          grepl(input$agency, Agency)
             )
             df(d)
             })

discrep_plot <- plot_ly(data = filter(good_data, !is.na(Discrepancy), grepl("Formal", Type))%>%
         group_by(Agency, Discrepancy)%>%
         summarize(count = n(), tot = n_distinct(NMFS.Tracking.Number))%>%
         group_by(Agency)%>%
         mutate(freq = count/sum(count))%>%
         filter(sum(tot) > 20)%>%
         bind_rows(filter(good_data, !is.na(Discrepancy), grepl("Formal", Type),
                          grepl("20[0-9][0-9]", Fiscal.Year))%>%
                     group_by(Discrepancy)%>%
                     summarize(count = n())%>%
                     mutate(freq = count/sum(count), Agency = "All Agencies")),
       type = 'bar', x = ~freq, y = ~Agency, color = ~Discrepancy,
       #text = ~count, textposition = 'outside',
       colors = c(plasma(12)[c(1,4,6)], 'grey', plasma(12)[c(8,10,12)]),
       hoverinfo = 'text',
       text = ~paste(Agency, count, round(freq, 2))
     )%>%
  layout(barmode = 'stack',
         legend = list(orientation = 'h', x = 0.3, y = 1.2, traceorder = "reversed", title = 'Discrepancy'),
         xaxis = list(title = "Proportion of Determinations",
                      titlefont = list(color = 'black'),
                      tickfont = list(color = 'black')),
         yaxis = list(title = "", tickfont = list(color = 'black', size = 10)),
         margin = list(l = 300, r = 0, t = 0))

#TimeLine
output$time_plot <- renderPlotly({
  plot_ly(data = filter(df(), grepl("20[0-16]", Fiscal.Year),
                        Type == "Formal Consultation"| Type == "Informal Consultation")%>%
                        group_by(NMFS.Tracking.Number)%>%
          summarize(FY = first(Fiscal.Year), Type = first(Type))%>%
          group_by(FY, Type)%>%
          summarize(count= n()),
        x = ~FY, y = ~count, color = ~ Type, type = 'bar', colors = rev(viridis(2)))%>%
  layout(barmode = 'stack',
         xaxis = list(title = "Fiscal Year"),
         yaxis = list(title = "Consultations"),
         legend = list(x = 0, y = 1.2),
         margin = list(b = 10))
})

#Frequency of Consultations by Work Category
cat_dt <- reactive({filter(df(), grepl(".", Category), grepl("20[0-9][0-9]", Fiscal.Year))%>%
  group_by(NMFS.Tracking.Number)%>%
  summarize(Category = first(Category))%>%
  group_by(Category)%>%
  summarize(count = n())%>%
  top_n(10, count)})

cat_names <- reactiveVal(c())

observeEvent(cat_dt(),
             {cat_vec <- isolate({cat_dt()$count})
             names(cat_vec) <- isolate({cat_dt()$Category})
             cat_names(names(sort(cat_vec)))
             })

output$category_plot <- renderPlotly({
  plot_ly(
    data = cat_dt(),
    y = ~Category,
    x = ~count,
    type= "bar",
    orientation = 'h',
    text = ~Category,
    textposition = 'outside',
    textfont = list(color = 'black'))%>%
  layout(title = "Work Type",
         yaxis = list(categoryorder = 'array',
                      categoryarray = cat_names(),
                      title = "",
                      showticklabels = FALSE),
         xaxis = list(title = "# of Consultations"
                      #range = c(0, 6000)
                      ),
         margin = list(r = 10, l = 10, b = 10, t = 100)
         )
})

#Frequency of Consultations by Agency
ag_dt <- reactive({filter(df(), grepl(".", Agency), grepl("20[0-9][0-9]", Fiscal.Year))%>%
  group_by(NMFS.Tracking.Number)%>%
  summarize(Agency = first(Agency))%>%
  group_by(Agency)%>%
  summarize(count = n())%>%
  top_n(10, count)})

ag_names <- reactiveVal(c())

observeEvent(ag_dt(),
             {ag_vec <- isolate({ag_dt()$count})
             names(ag_vec) <- isolate({ag_dt()$Agency})
             ag_names(names(sort(ag_vec)))
             })

# ag_vec <- ag_dt$count
# names(ag_vec) <- ag_dt$Agency

output$agency_plot <- renderPlotly({
  plot_ly(data = ag_dt(), y = ~Agency, x = ~count, type= "bar", orientation = 'h',
        text = ~Agency,
        textposition = 'outside',
        textfont = list(color = 'black'))%>%
  layout(title = "Federal Agencies",
         yaxis = list(categoryorder = 'array',
                      categoryarray = ag_names(),
                      title = "",
                      showticklabels = FALSE),
         xaxis = list(title = "# of Consultations"),
         margin = list(b = 10, t = 100, l = 10, r = 10)
  )
})

#  Frequency of Consultations by Species
sp_dt <- reactive({filter(df(), grepl(".", Common.Name), grepl("20[0-9][0-9]", Fiscal.Year))%>%
  group_by(Common.Name)%>%
  summarize(count = length(unique(NMFS.Tracking.Number)))%>%
  top_n(10, count)})

sp_names <- reactiveVal(c())

observeEvent(sp_dt(),
             {sp_vec <- isolate({sp_dt()$count})
             names(sp_vec) <- isolate({sp_dt()$Common.Name})
             sp_names(names(sort(sp_vec)))
             })

# sp_vec <- sp_dt$count
# names(sp_vec) <- sp_dt$Common.Name

output$species_plot <- renderPlotly({
  plot_ly(data = sp_dt(), y = ~Common.Name, x = ~count, type= "bar", orientation = 'h',
        text = ~Common.Name,
        textposition = 'outside',
        textfont = list(color = 'black', size = 14)
        #textposition = c('inside', rep('outside', 3), 'inside', rep('outside', 5)),
        #textfont = list(color = c('white', rep('black', 3), 'white', rep('black', 5)), size = 14)
        )%>%
  layout(title = "Species",
         titlefont = list(color = 'black', size = 16),
         yaxis = list(categoryorder = 'array',
                      categoryarray = sp_names(),
                      title = "",
                      showticklabels = FALSE),
         xaxis = list(title = "# Consultations",
                      #range = c(0, 15000),
                      showgrid = FALSE,
                      tickfont = list(color = 'black', size = 12)),
         margin = list(b = 10, t = 100, l = 10, r = 10)
  )
})

###CREATE MATRICES AND PERMUTATION TESTS FOR SPECIES BY CATEGORY HEATMAP
permtest <- permatfull(species_x_category, fixedmar = 'both', mtype = "count", times = 1000)
permtests <- lapply(permtest$perm, function(i){return(i >= species_x_category)})
permeff <- species_x_category - Reduce('+', permtest$perm)/1000
permstats <- Reduce('+', permtests)/1000


spbycat_plot <- plot_ly(z = t(permeff), #species_x_category
            y = ~colnames(species_x_category),
            x = ~rownames(species_x_category), type = 'heatmap',
            zmin = -5, zmax = 10
            )%>%
      layout(margin = list(b = 100, l = 10),
             xaxis = list(title = "",
                          tickfont = list(color = 'black')),
             yaxis = list(title = "",
                          tickfont = list(color = 'black')))


```

### The U.S. Endangered Species Act (ESA) requires that federal agencies consult with the U.S. Fish and Wildlife Service (FWS) or National Marine Fisheries Service (NMFS) to ensure federal actions do not jeopardize the existence of listed species. We analyzed all the consultations recorded by NMFS from 2000–2017 and to investigate the role of expert review and analysis in the consultation process.

```{r intro, echo = FALSE}
fluidPage(
  fluidRow(
    column(5,
            p("In a recent [paper](https://www.biorxiv.org/content/10.1101/424911v1.article-info), we provide a comprehensive assessment of how a key provision of the U.S. Endangered Species Act (ESA) is implemented: consultation to ensure federal actions do not jeopardize the existence of listed species. Data from all 24,893 consultations recorded by the National Marine Fisheries Service (NMFS) from 2000–2017 show federal agencies and NMFS frequently agreed (79%) on how federal actions would affect listed species. In cases of disagreement agencies most often (71%) underestimated effects relative to the conclusions of species experts at NMFS. Such instances can have deleterious consequences for imperiled species. In 22 consultations covering 14 species, agencies concluded that an action would not harm species while NMFS determined the action would jeopardize species’ existence. These results affirm the importance of NMFS role in preventing federal actions from jeopardizing listed species. Excluding expert agencies from consultation compromises biodiversity conservation, but we identify approaches that improve consultation efficiency without sacrificing species protections.")
            ),
    column(7,
           img(src = 'Box1.png', width = 400, height = 400,
               alt = 'Conceptual diagram of the section 7 consultation process under the Endangered Species Act')
    )
  )
)
```

### Agreement among Agencies

```{r discrep, echo = FALSE}
fluidPage(
    discrep_plot
)
```

***

When formal consultation is initiated, federal agencies propose a determination as to how their planned actions will affect listed species. During consultation NMFS may agree or disagree with these proposals.  We scored the degree of agreement on an ordinal -3 to 3 scale, with -3 representing the most severe underestimation of effects by agencies possible, and 3 representing the most severe overestimation of effects by agencies relative to NMFS conclusion

### General Patterns

```{r selectors, echo = FALSE}
fluidPage(
  fluidRow(
    column(3,
           select_taxa,
           select_category,
           select_agency,
           run_update
           ),
    column(9,
           plotlyOutput("time_plot")
           )
    ),
  br(),
  fluidRow(
    column(4,
           plotlyOutput("agency_plot")
           ),
    column(4,
           plotlyOutput("category_plot")
           ),
    column(4,
           plotlyOutput("species_plot")
    )
  )
)

```
